---
output: github_document
title: "Routing with stplanr"
author: "Robin Lovelace"
date: "`r Sys.Date()`"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

As described in the `stplanr-paper` vignette there are various approaches to routing available in the package.
This vignette seeks to demonstrate these approaches using a case study in the town of Jena, Germany.

The prerequisites are as follows:

```{r}
library(stplanr)
library(tmap)
library(sf)
library(leaflet)
library(dodgr)
```


# Online routing

The simplest routing strategy is to use an online service such as the public facing OSRM demo routing service at http://router.project-osrm.org/ with an interactive map-based GUI at http://map.project-osrm.org/.

We can demonstrate this routing service in action by planning a route from Jena Paradies rail station to Jen Tower, a building that is "the tallest in Eastern Germany"  according to [Wikipedia](https://en.wikipedia.org/wiki/JenTower).
This represents a typical journey that one may take on visiting the city for the first time:

```{r}
from = "Jena Paradies"
to = "JenTower"
```

```{r, eval=FALSE}
r_osrm = route_osrm(from, to)
saveRDS(r_osrm, "r_osrm.Rds")
```

What just happened?
There were 3 main steps taken inside the function `route_osrm()`:

- The origin and destination text strings were geo-coded
- The information was sent to the OSRM server and returned
- The data was converted into a form that can be used in R

In case this does not work on your computer (you need a recent version of stplanr) you can load it as follows:

```{r}
r_osrm = readRDS("r_osrm.Rds")
```


The class of the resulting object can be queried as follows:

```{r}
class(r_osrm)
```

We can verify that it contains the spatial information we need by plotting it interactively:

```{r}
leaflet() %>% addTiles() %>% addPolylines(data = r_osrm)
```

We can execute the first step taken as follows:

```{r}
(from_coords = geo_code(from))
(to_coords = geo_code(to))
```

We could continue to explore what happened in the function by executing the functions `viaroute()` and `viaroute2sldf()` sequentially but that is not the point of this vignette (see the source code for more information on the internals of `route_osrm` but be warned, they are not pretty)!

Other routing services can also be used, as illustrated by the `route_graphhopper()` call below:

```{r, eval=FALSE}
r_graphbike = route_graphhopper(from_coords, to_coords)
saveRDS(r_graphbike, "r_graphbike.Rds")
```

Be warned: this code may not work on your computer because you need a graphhopper api key.
We can check that we have one on this computer with the following command:

```{r}
Sys.getenv("GRAPHHOPPER") %>% nchar()
```

If so you will have 32 characters, if not the result will have 0.
(To get a Graphhopper API key see `?route_graphhopper`.)
That is not a problem as I've saved the result, which can be loaded with:

```{r}
r_graphbike = readRDS("r_graphbike.Rds")
```

The advantage of the Graphhopper API is that it has three routing profiles available (and many more with their paid plan), as illustrated below (the default is bike):

```{r, eval=FALSE}
r_graphcar = route_graphhopper(from_coords, to_coords, vehicle = "car")
r_graphwalk = route_graphhopper(from_coords, to_coords, vehicle = "foot")
saveRDS(r_graphcar, "r_graphcar.Rds")
saveRDS(r_graphwalk, "r_graphwalk.Rds")
```

As before, these can be loaded from the repo as follows:

```{r}
r_graphcar = readRDS("r_graphcar.Rds")
r_graphwalk = readRDS("r_graphwalk.Rds")
```


We can compare these different routes, as illustrated in the figures below:

```{r routes, fig.show='hold', fig.cap="Routes for driving (left panel, with OSRM in red and Graphhopper in orange) and active transport (right panel, cycling in green, walking in blue).", out.width="50%"}
b = tmaptools::bb(r_osrm, ext = 1.2)
tm_shape(r_osrm, bbox = b) +
  tm_lines(col = "red", lwd = 6) +
  qtm(r_graphcar, lines.col = "orange", lines.lwd = 4)
tm_shape(r_osrm, bbox = b) +
  tm_lines(col = "white", lwd = 6) +
  qtm(r_graphbike, lines.col = "green", lines.lwd = 3) +
  qtm(r_graphwalk, lines.col = "blue", lines.lwd = 2)
```

The results show a diversity of route options.
We can make the following generalisations:

- Motorised routes tend to be more circuitous
- Routes for the same mode can be very different depending on which weighting profile is used
- Walking and cycling routes tend to be similar, although walking routes tend to be more direct as they can pass through steps etc.

There are some major limitations associated with online routing:

- It is slow
- It can be expensive
- It's not conducive to reproducibility - the service may change and you need an API key 
- You do not have control over the weighting profiles (this can be good and bad)

To overcome these limitations we can do local routing.

# Local routing

To do local routing you need a route network, e.g. as provided by the following commands:

```{r, eval=FALSE}
library(osmdata)
bb_jena = getbb("Jena")
osm_jena = opq(bbox = bb_jena) %>% 
  add_osm_feature("highway", "prim|second|cycle", value_exact = FALSE) %>% 
  osmdata_sf()
summary(osm_jena)
ways = osm_jena$osm_lines
write_sf(ways, "ways.geojson")
```

To save time we've saved the result, which can be loaded as follows from the vignettes folder:

```{r}
ways = read_sf("ways.geojson")
```


Let's check if this is a reasonable representation of Jena's route network:

```{r}
leaflet() %>% addTiles() %>% addPolylines(data = ways)
```

Clearly it's a very simplified route network.
A more comprehensive network could be created by altering the arguments passed to `add_osm_feature`, e.g. to simply `add_osm_feature("highway")`.
We deliberately use a subset of the network for teaching.
Now, how do we find routes along it?

```{r}

```


